A Type System for Safe and Scalable Parallelism using Logical Regions

### 2012-07-05 doc/popl2013/langdef.tex
Adding some labels to the langdef
checkpoint

modify $M$
\begin{table*}
\centering
{\small
\begin{tabular}{l|p{5cm}|p{5cm}|p{5cm}}
{\bf Rule} & 
\multicolumn{1}{|c|}{$\mathbf{v \sim_H M \llbracket T \rrbracket}$} &
\multicolumn{1}{|c|}{$\mathbf{E \sim H}$} &
\multicolumn{1}{|c|}{$\mathbf{E : \Phi}$} \\
\hline
BoolConst & & & \\
IntConst & & & \\
TupleBuild & & & \\
TupleField & & & \\
LocalVar & & & \\
NullPtr & & & \\
NewPtr & & & \\
IsNull & & & \\
UpRegion & & & \\
DownRegion & & & \\
Read & & & \\
Write & & & \\
Reduce & & & \\
Newcolor & & & \\
Color & & & \\
IntMath & & & \\
IntCmp & & & \\
Let & & & \\
IfElse & & & \\
TaskCall & & & \\
Partition & & & \\
Pack & & & \\
Unpack & & &
\end{tabular}}
\caption{Proof pieces}
\end{table*}

@@

\begin{thm}
Let $S_0$ be an initial store, $C$ be a clobber set and $\oton{E}$ be memory traces to be
interleaved.  Let $E' = E_1 \dplus \ldots \dplus E_n$.  If there exists any $E''$ that is a
valid interleaving of $\oton{E}$, then $E'$ is also valid interleaving of $\oton{E}$.
\end{thm}

@@

\begin{lem}
Let $S$ be an initial store, $C$ be a clobber set, $\oton{E}, E'_a, E'_b$ be memory traces, and
$\epsilon_1, \epsilon_2$ be two annotated memory operations such that $\epsilon_1 \# \epsilon_2$.
Then,
$$valid\_interleave(S, C, E'_a \dplus \epsilon_1 \dplus \epsilon_2 \dplus E'_b, \oton{E})$$
if and only if
$$valid\_interleave(S, C, E'_a \dplus \epsilon_2 \dplus \epsilon_1 \dplus E'_b, \oton{E}).$$
\end{lem}

@@

First, we would like a fallback approach that will always result in a valid interleaving, even if
we lack information about the constituent traces (or perhaps lack the time to analyze it).  The

The soundness proof above depended, in part, on the properties of a valid interleaving of two
memory traces.  We now consider algorithms that create interleavings that satisfy those
conditions.

The simplest way to achieve a valid interleaving is to concatenate the traces in order.  This
will always result in a valid interleaving, but implies a serialization of the evaluation of
subexpressions.

@@

\framebox{$\forall S, C, E_1, E_2. valid\_interleave(S, C, E_1 \dplus E_2, E_1, E_2)$}

Next step: For noninterfering traces $E_1 \# E_2$, show any interleaving is valid.

Third step: For interactions only on atomic-coherence locations, serializability is sufficient.

For a given expression $e$, we can consider $v^o$ to be a ``sequential'' result and $E^o$ to be the corresponding ``sequential'' effects if: \\

\framebox{$M, L, H, S, \emptyset \vdash e \mapsto v^o, E^o$} \\

(i.e. if there are no manipulations of the store other than what's done by
the evaluation of the expression $e$.)  We now wish to consider what parallel
evaluations of this expression with other expression evaluations will still
yield our ``sequential'' result. \\

Two effects are considered to be noninterfering, written $E_1 \# E_2$ if for
every $\epsilon_1$ in $E_1$ and every $\epsilon_2$ in $E_2$,
\begin{enumerate}
\item $\epsilon_1$ and $\epsilon_2$ are to different memory locations, or
\item $\epsilon_1 = read(l, v_1)$ and $\epsilon_2 = read(l, v_2)$, or
\item $\epsilon_1 = reduce_{id}(l, v_1)$ and $\epsilon_2 = reduce_{id}(l, v_2)$
\end{enumerate}

We define nonintereference (under a given mapping $M$) of static privileges
$\Phi_1$ and $\Phi_2$ similarly.  We use the following table for determining
noninterference of two individual priviledges $\phi_1$ and $\phi_2$:

\begin{tabular}{ccccc}
\multicolumn{2}{l}{$\phi_1 \#_M \phi_2 ?$} & \multicolumn{3}{c}{$\phi_2$} \\
\multicolumn{2}{l}{} & $reads(r_2)$ & $writes(r_2)$ & $reduces(id_2, r_2)$ \\
\multirow{3}{*}{$\phi_1$} & $reads(r_1)$ & true & $M(r_1) \cap M(r_2) = \emptyset$ & $M(r_1) \cap M(r_2) = \emptyset$\\
& $writes(r_1)$ & $M(r_1) \cap M(r_2) = \emptyset$ & $M(r_1) \cap M(r_2) = \emptyset$ & $M(r_1) \cap M(r_2) = \emptyset$ \\
& $reduces(id_1, r_1)$ & $M(r_1) \cap M(r_2) = \emptyset$ & $M(r_1) \cap M(r_2) = \emptyset$ & $\begin{array}{l}id_1 = id_2 \vee \\ M(r_1) \cap M(r_2) = \emptyset\end{array}$
\end{tabular}

We can then define: \\

\framebox{$\Phi_1 \#_M \Phi_2 \Leftrightarrow \displaystyle\bigwedge_{\begin{array}{c}\phi_1 \in \Phi_1 \\ \phi_2 \in \Phi_2 \end{array}} \phi_1 \#_M \phi_2$} \\

From the soundness of effects, result, we can show that for two expressions
$e_1$ and $e_2$, if their privileges are noninterfering, any potential memory
traces from their evaluation must also be noninterfering.  Formally:

\finfrule
{\begin{array}{l}
\typeenvx[P=\Phi_1] e_1 : T_1 \\
\typeenvx[P=\Phi_2] e_2 : T_2 \\
\opsenvx e_1 \mapsto v_1, E_1 \\
\opsenvx e_2 \mapsto v_2, E_2 \\
\Phi_1 \#_M \Phi_2
\end{array}}
{E_1 \# E_2}

(TODO: Generalize this further and allow $M_1 \not= M_2$?) \\

Note that this result holds even if $C \not= \emptyset$.  This allows 
two ``locally independent'' subtasks to run in parallel even if one or both
of them is interacting (in a programmer-permitted way) with a third subtask.

(TODO: Talk about coherence modes and how the runtime orders subtasks.  Maybe
describe it in terms of a ``canonical'' memory trace and then allow
reorderings?)

>>>>

%% \begin{table*}
%% \centering
%% {\small
%% \begin{tabular}{l|p{5cm}|p{5cm}|p{5cm}}
%% {\bf Rule} & 
%% \multicolumn{1}{|c|}{$\mathbf{v \sim_H M \llbracket T \rrbracket}$} &
%% \multicolumn{1}{|c|}{$\mathbf{E \sim H}$} &
%% \multicolumn{1}{|c|}{$\mathbf{E : \Phi}$} \\
%% \hline
%% BoolConst & & & \\
%% IntConst & & & \\
%% TupleBuild & & & \\
%% TupleField & & & \\
%% LocalVar & & & \\
%% NullPtr & & & \\
%% NewPtr & & & \\
%% IsNull & & & \\
%% UpRegion & & & \\
%% DownRegion & & & \\
%% Read & & & \\
%% Write & & & \\
%% Reduce & & & \\
%% Newcolor & & & \\
%% Color & & & \\
%% IntMath & & & \\
%% IntCmp & & & \\
%% Let & & & \\
%% IfElse & & & \\
%% TaskCall & & & \\
%% Partition & & & \\
%% Pack & & & \\
%% Unpack & & &
%% \end{tabular}}
%% \caption{Proof pieces}
%% \end{table*}

@@

\begin{thm}
\label{thm:sequential}
Let $S_0$ be an initial store, $C$ be a clobber set and $\oton{E}$ be memory traces to be
interleaved.  Let $E' = E_1 \dplus \ldots \dplus E_n$.  If there exists any $E''$ that is a
valid interleaving of $\oton{E}$, then $E'$ is also valid interleaving of $\oton{E}$.
\end{thm}

@@

\begin{lem}
\label{lem:nonintswap}
Let $S$ be an initial store, $C$ be a clobber set, $\oton{E}, E'_a, E'_b$ be memory traces, and
$\epsilon_1, \epsilon_2$ be two annotated memory operations such that $\epsilon_1 \# \epsilon_2$.
Then,
$$valid\_interleave(S, C, E'_a \dplus \epsilon_1 \dplus \epsilon_2 \dplus E'_b, \oton{E})$$
if and only if
$$valid\_interleave(S, C, E'_a \dplus \epsilon_2 \dplus \epsilon_1 \dplus E'_b, \oton{E}).$$
\end{lem}

@@

We now extend the noninterference operator to entire memory traces, providing a way to say that
no operation in a trace interferes with any operation in other:
$$E_1 \# E_2 \Leftrightarrow \displaystyle\bigwedge_{\epsilon_1 \text{ in } E_1, \epsilon_2 \text{ in } E_2} \epsilon_1 \# \epsilon_2$$
and observe that this property permits arbitrary ordering (i.e. parallel execution) of the 
corresponding subexpressions.

\begin{lem}
\label{lem:noninteffects}
Let $S$ be an initial store, $C$ be a clobber set, $\oton{E}$ be memory traces such that $E_i \# E_j$
for every $1 \leq i < j \leq n$.  Then any trace $E'$ satisfying:
$$any\_interleave(E', \oton{E})$$
is a valid interleaving.
\end{lem}

\begin{proof}
We will use Theorem~\ref{thm:sequential} and Lemma~\ref{lem:nonintswap} to make repeated swaps
of adjacent operations to transform $E'$ into $E_1 \dplus \ldots \dplus E_n$.  The symmetry of
Lemma~\ref{lem:nonintswap} is then used to show that the validity of $E_1 \dplus \ldots \dplus E_n$
under Theorem~\ref{thm:sequential} implies the validity of $E'$.

The swapping algorithm as follows: Associate with each operation in $E'$ the index of that operation
in $E_1 \dplus \ldots \dplus E_n$.  Any consectutive pair of operations $\epsilon_1$ and $\epsilon_2$
for which $index(\epsilon_1) > index(\epsilon_2)$ is a {\em misordered pair}.  For any such pair,
$\epsilon_1$ and $\epsilon_2$ will be from different constituent traces (if they were from the same
trace, their misordering would mean that $E'$ was not an interleaving of $\oton{E}$).  Every pair of
traces is noninterfering, so $\epsilon_1 \# \epsilon_2$ and they can be swapped while preserving
validity of the trace.  At most $n^2/2$ swaps are needed to eliminate all misordered pairs.
\end{proof}

Our next step is to extend noninterference to the privileges used in the type system.  We will make
use of the region mapping $M$ to translate the logical regions named in privileges into sets of
locations (i.e. physical regions), and define noninterference between individual privileges as
follows:
\begin{center}
\begin{tabular}{llc}
\multicolumn{1}{c}{$\phi_1$} & \multicolumn{1}{c}{$\phi_2$} & $\phi_1 \#_M \phi_2$? \\
\multirow{3}{*}{$reads(r_1)$} & $reads(r_2)$ & true \\
& $writes(r_2)$ & $M(r_1) \cap M(r_2) = \emptyset$ \\
& $reduces(id_2,r_2)$ & $M(r_1) \cap M(r_2) = \emptyset$ \\
\cline{1-1}
\multirow{3}{*}{$writes(r_1)$} & $reads(r_2)$ & $M(r_1) \cap M(r_2) = \emptyset$ \\
& $writes(r_2)$ & $M(r_1) \cap M(r_2) = \emptyset$ \\
& $reduces(id_2,r_2)$ & $M(r_1) \cap M(r_2) = \emptyset$ \\
\cline{1-1}
\multirow{3}{*}{$reduces(id_1r_1)$} & $reads(r_2)$ & $M(r_1) \cap M(r_2) = \emptyset$ \\
& $writes(r_2)$ & $M(r_1) \cap M(r_2) = \emptyset$ \\
& $reduces(id_2,r_2)$ & $\begin{array}{c}id_1 = id_2 \vee \\ M(r_1) \cap M(r_2) = \emptyset\end{array}$ \\
\end{tabular}
\end{center}
and noninterference between sets of privileges in the same way as memory traces:
$$\Phi_1 \#_M \Phi_2 \Leftrightarrow \displaystyle\bigwedge_{\phi_1 \in \Phi_1, \phi_2 \in \Phi_2} \phi_1 \#_M \phi_2$$

\begin{lem}
\label{lem:nonintpriv}
Let $M$ be a region mapping and $E_1$ and $E_2$ two memory traces such that $M \vdash E_1 : \Phi_1$ and $M \vdash E_2 : \Phi_2$.  If $\Phi_1$ and $\Phi_2$ are noninterfering under $M$, then $E_1$ and $E_2$ must be noninterfering.
\end{lem}

\begin{proof}
The proof is by contradiction.  Assume $E_1 \cancel\# E_2$.  Then there must be some $\epsilon_1$ in $E_1$ and some $\epsilon_2$ in $E_2$ such that $\epsilon_1 \cancel\# \epsilon_2$.  This requires that
$\epsilon_1$ and $\epsilon_2$ be operations to the same location $l$ and not both be reads or both
reductions using the same function name.  Assume $\epsilon_1 = read(l, c_1, v_1)$ and $\epsilon_2 = write(l, c_2, v_2)$.  The definition of $M \vdash E_1 : \Phi_1$ tells us that there must be some $r_1$ satisfying $l \in M(r_1) \wedge reads(r_1) \in \Phi_1$.  Similarly, there must be some $r_2$ satisfying $l \in M(r_2) \wedge writes(r_2) \in \Phi_2$.  Letting $\phi_1 = reads(r_1)$ and $\phi_2 = writes(r_2)$, we have $M(r_1) \cap M(r_2) \not= \emptyset$ (it contains at least $l$), so $\phi_1 \cancel\#_M \phi_2$ and $\Phi_1 \cancel\#_M \Phi_2$.  There are seven other cases to consider (using $id_1 \not= id_2$ for the reduce-reduce case), but all yield the same result.
\end{proof}

We are now able to connect the dots and get to the key result that allows the Legion runtime to 
safely run subtasks in parallel.

\begin{thm}
Let $\oton{e}$ be well-typed Legion expressions, each with its own privileges $\Phi_i$.
Let $M$ be a region mapping, $L$ be local value mapping,
$H$ be a heap typing, and $S$ be an initial state of the store satisfying the consistency properties:
$$
\begin{array}{l}
M \sim \Omega \\
L \sim_H \Gamma \\
S \sim H
\end{array}
$$
If $\Phi_i \#_M \Phi_j$ for $1 \leq i < j \leq n$, then any parallel execution of expressions
$\oton{e}$ will result in a valid interleaving of memory operations.
\end{thm}

The proof follows directly from Lemma~\ref{lem:nonintpriv} and Lemma~\ref{lem:noninteffects}. \\

Note that this result holds even if the clobber set $C$ is non-empty.  This allows 
two ``locally independent'' subtasks to run in parallel even if one or both
of them is interacting (in a programmer-permitted way) with a third subtask.

TODO: talk about atomic case

@@
